Explainability and Adversarial Robustness for RNNs

### 2019-12-20 paper/paper.tex
Streamlined to US english

Based on the insights gained by the feature importance and explainability methods, we propose two defence methods.
%develop a simple method for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process.
First, by simply leaving out manipulable features, we obtain a classifier which is slightly less accurate but is still useful to be deployed in a real scenario. Second,
by deploying an adversarial training procedure,
%AH. compared to other research about adversarial training, I dont feel that out approach is so special
%develop a custom adversarial training procedure which especially focuses on ease of use and low resource consumption and aims to keep the number of additionally introduced training hyperparameters low. With this novel strategy
we can reduce the attack surface and harden the resulting \gls{ids} while retaining all features and similar classification performance. % like for the original \gls{ml} model.

@@

We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e. forward or reverse path) and all TCP flags (0 if the flow is not TCP).
% \begin{itemize*}
% \item source port
% \item destination port
% \item protocol identifier
% \item packet sizes
% \item \glspl{iat} of packets
% \item packet directions (if packet goes there or back)
% \item all TCP flags (0 if the flow is not TCP)
% \end{itemize*}.
We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
We use Z-score normalization and a train/test split of 2:1.

@@

\subsection{$L_\infty$-bounded Projected Gradient Descent}
\cite{madry_towards_2018} uses a method for generating adversarial samples which deploys \gls{pgd}  to maximize the network's loss function while constraining the achieved $L_\infty$ distance from the original samples. Hence, an important difference to \gls{cw} is that for this method the network's loss instead of its logit output is considered.
We consider this method an interesting combination of \gls{cw} and \gls{fgsm}.
Since its functioning is somewhat similar to \gls{cw}, we expected the generated adversarial samples to be similarly close the the original samples.

@@

\section{Defences}
We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks.
\subsection{Reducing Features}
The most obvious defence, which is only possible because of the nature of the dataset, is to simply omitting features an attacker can manipulate. We try two different approaches:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Leaving out all manipulable features: packet size and \gls{iat}
\item Leaving out manipulable features only when the attacker can manipulate them, i.e. only in the direction from the attacker to the victim. This, however, does not prevent adversarial samples for botnets, for which both sides are malicious.
\end{itemize}

>>>>

Based on the insights gained by the feature importance and explainability methods, we propose two defense methods.
%develop a simple method for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process.
First, by simply leaving out manipulable features, we obtain a classifier which is slightly less accurate but is still useful to be deployed in a real scenario. Second,
by deploying an adversarial training procedure,
%AH. compared to other research about adversarial training, I dont feel that out approach is so special
%develop a custom adversarial training procedure which especially focuses on ease of use and low resource consumption and aims to keep the number of additionally introduced training hyperparameters low. With this novel strategy
we can reduce the attack surface and harden the resulting \gls{ids} while retaining all features and similar classification performance. % like for the original \gls{ml} model.

@@

We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e. forward or reverse path) and all TCP flags (0 if the flow is not TCP).
% \begin{itemize*}
% \item source port
% \item destination port
% \item protocol identifier
% \item packet sizes
% \item \glspl{iat} of packets
% \item packet directions (if packet goes there or back)
% \item all TCP flags (0 if the flow is not TCP)
% \end{itemize*}.
We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behavior \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
We use Z-score normalization and a train/test split of 2:1.

@@

\subsection{$L_\infty$-bounded Projected Gradient Descent}
\cite{madry_towards_2018} uses a method for generating adversarial samples which deploys \gls{pgd}  to maximize the network's loss function while constraining the achieved $L_\infty$ distance from the original samples. Hence, an important difference to \gls{cw} is that for this method the network's loss instead of its logit output is considered.
We consider this method an interesting combination of \gls{cw} and \gls{fgsm}.
Since its functioning is somewhat similar to \gls{cw}, we expected the generated adversarial samples to be similarly close to the original samples.

@@

\section{Defenses}
We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks.
\subsection{Reducing Features}
The most obvious defense, which is only possible because of the nature of the dataset, is to simply omitting features an attacker can manipulate. We try two different approaches:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Leaving out all manipulable features: packet size and \gls{iat}
\item Leaving out manipulable features only when the attacker can manipulate them, i.e. only in the direction from the attacker to the victim. This, however, does not prevent adversarial samples for botnets, for which both sides are malicious.
\end{itemize}
